---
title: "ENVS193DS-code"
format: html
editor: visual
author: "Valentina Caro"
date: "03-13-2026"
execute: 
  message: false
  warning: false
---

## Setup

```{r}
#| label: reading in packages
#| message: false

library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # organization
library(readxl) # reading excel
library(scales) # modifying axis labels
library(ggeffects) # model predictions
library(MuMIn) # model selection
library(lterdatasampler) # dataset from LTER
library(extrafont) # adds more fonts
library(showtext) # allow changing font
library(flextable) # add flextable
library(effectsize) # calculating effect sizes

kelp <- read.csv(
  here("data",
       "Annual_Kelp_All_Years_20250903.csv")
) # creating kelp dataframe

mopl <- read.csv(
  here("data",
       "MOPL_nest-site_selection_Duchardt_et_al._2019.csv")
)

apple <- read_xlsx(
  here("data",
       "apple.xlsx")
)

```

HTML Link: <https://valecarov.github.io/ENVS-193DS_winter_2026_final/code/193ds-code.html>

# Problem 1

## a) Transparent Statistical Methods

In part 1, my coworkers used a linear regression to test whether precipitation predicts flooded wetland area. In part 2, my coworkers used a Kruskal-Wallis test to determine whether the median flooded wetland area differs across 5 water year classifications.

## b) Figure Needed

An appropriate figure to accompany part 2 of the report would be a box and jitter plot to show the median and overall spread across the water year classifications. The x-axis/group should be water year classification (wet, above normal, below normal, dry, and critical drought), and the y axis should be flooded wetland area in m\^2. Each classification should have its own custom color to ensure data is clearly displayed.

## c) Suggestions for Rewriting

Based on our model predictions, we found that precipitation does not predict flooded wetland area (m\^2)(Linear Regression, F(1, df) = f-statisic, R^2^ = R\^2 value, p = 0.11, $\alpha$ = significance level). We did not find a difference ( $\eta^2$ = effect size) in median flooded wetland area (m^2)^ between water year classifications (wet, above normal, below normal, cry, critical drought) (Kruskal-Wallis test, $\chi^2$ (df) = H-statistic, p = 0.12, $\alpha$ = significance level).

## d) Interpretation

One additional variable that could impact the flooded wetland area of the San Joaquin River Delta is water management. As there is not sufficient evidence of precipitation having an effect on flooded wetland area in this study, it is likely that the management of the water is a stronger predictor than the amount of water itself. This could be tested by measuring the flooded wetland area under different water management strategies to determine which strategies may increase flooded wetland area, maximizing waterfowl breeding success.

# Problem 2

## a) Cleaning and Summarizing

```{r}
#| label: cleaning & summarizing

kelp_clean <- kelp |> 
  clean_names() |> # clean the names
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> # filter for sites IVEE, NAPL, & CARP
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> # exclude specific date, transect, and quad values
  mutate(site_name = case_when( # create column for site names from site, using full name rather than abbreviation
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) |> 
  mutate(site_name = as_factor(site_name),
       site_name = fct_relevel(site_name,
                               "Naples", "Isla Vista", "Carpinteria")) |> # make site name a factor and set order
  group_by(site_name, year) |> # group by site name and year
  summarize(mean_fronds = mean(fronds)) |> # calculate the mean frond size
  ungroup() # undo group_by function

```

```{r}
#| label: check structure

str(kelp_clean) # check structure of the data frame

```

```{r}
#| label: display

slice_sample(kelp_clean,
             n = 5) # display 5 random rows from data frame

```

## b) Understanding the Data

Data from 09-22-2000 in transect 6 in quad 40 was excluded from kelp_clean due to a missing data sheet causing invalid values for the number of fronds and the diameter of the holdfast. The number and size (number of fronds and diameter of the holdfast) of kelp in this line of the table are both -99999. (Abstract section under Summary tab of data page)

## c) Visualizing the Data

```{r}
#| label: add font

font_add(family = "Times New Roman", regular = "times.ttf") # adds Times New Roman font
showtext_auto() # allows font to be used 

```

```{r}
#| label: line graph

ggplot(data = kelp_clean,# use kelp_clean data frame
       aes(x = year, # set x and y axis
           y = mean_fronds,
           color = site_name)) +
  geom_line() + # adding line graph
  geom_point() + # adding points
  labs(title= "Sites vary in mean kelp fronds per year", # setting titles
       x = "Year",
       y = "Mean kelp fronds per year",
       color = "Site") +
  scale_color_manual(values = c( # set custom colors
    "Naples" = "darksalmon",
    "Isla Vista" = "palevioletred",
    "Carpinteria" = "deeppink4"
  )) +
  theme_minimal() + # change theme 
  theme(plot.title = element_text(hjust = 0.5)) + # move plot title to center
  theme(text = element_text(family = "Times New Roman", # change font, color, and size
                            color = "#1D0200",
                            size = 16)) +
  theme(axis.text = element_text(color = "#1D0200")) # change axis color to match text color
  
```

## d) Interpretation

The site with the most variability in mean kelp fronds per year was the Naples site, with values as low as about 4 fronds up to nearly 40 fronds, visualized by the huge spikes in the graph above. The Carpinteria site had the least variability in mean kelp fronds per year, ranging from about 3 fronds to 17 fronds as seen by the much smaller spikes.

# Problem 3

## a) Response Variable

In this dataset, the 0s and 1s in the binshrub and binsubshrub columns indicate whether there is shrub or sub-shrub cover in a nest-site. A 0 indicates no shrub or sub-shrub cover and a 1 indicates there is shrub or sub-shrub cover in the nest-site.

## b) Purpose of Study

### Visual Obstruction:

Visual obstruction is a continuous variable measured at the nest, and at 5 m and 10 m from the nest. As visual obstruction increases, the probability of Mountain Plover nest site use would decrease because mountain plovers have a positive response to sites with a higher percentage of bare ground (the least visual obstruction) and a negative response to increasing maximum vegetation height, and presence of shrub cover and cactus cover (higher visual obstruction) according to the nest-selection subsection under the results section.

### Distance to Prairie Dog Colony Edge:

Distance to prairie dog colony edge is a continuous variable. As distance to the colony edge increases, the probability of a Mountain Plover nest site would decrease. According to the 2nd and 3rd paragraphs in the Discussion section of the paper, this may be due to complementary resources outside of the colony. It may be easier for adult plovers to travel with chicks if they are closer to the edge of the colony rather than at the center. Additionally, mountain plover density was highest in colonies that were 100-500 ha in size and low in larger colonies according to Figure 2 in the Results section, which could be related to distance from the edge as plovers would have to travel much further distances in very large colonies. I believe mountain plovers prefer nest sites that offer protection while still allowing for easy travel to outside resources, causing density to be the highest near the edges of moderately sized colonies.

## c) Table of Models

\| Model Number \| Visual Obstruction \| Distance to Prairie Dog Colony \| Model Description \|

\|:-----------------:\|:---------------------:\|:----------------------------------:\|:--------------------:\|

\|:--------0--------:\|:---------------------:\|:----------------------------------:\|:--------null--------:\|

\|:--------1--------:\|:---------x-----------:\|:----------------x-----------------:\|:--------all---------:\|

\|:--------2--------:\|:---------x-----------:\|:----------------------------------:\|:--------vo---------:\|

\|:--------3--------:\|:---------------------:\|:----------------x-----------------:\|:-----edgedist-----:\|

## d) Run the Models

```{r}
#| label: cleaning

mopl_clean <- mopl |> # create new dataframe
  clean_names() |> # clean names
  mutate( used_bin = case_when( # create new column used_bin from id column
    id == "used" ~ "1",
    id == "available" ~ "0"
  )) |> 
  select(used_bin, edgedist, vo_nest) # select variables

```

```{r}
#| label: run models

# model 0: null model
model0 <- lm(used_bin ~ 1,
             data = mopl_clean)
# model 1: all predictors
model1 <- lm(used_bin ~ vo_nest + edgedist,
             data = mopl_clean)
# model 2: visual obstruction
model2 <- lm(used_bin ~ vo_nest,
             data = mopl_clean)
# model 3: distance to edge of colony
model3 <- lm(used_bin ~ edgedist,
             data = mopl_clean)

```

## e) Select the Best Model

```{r}
#| label: aic

AICc(model0,
     model1,
     model2,
     model3) |> 
  arrange(AICc)

```

The best model as determined by Akaike’s Information Criterion (AIC) is the model that only includes visual obstruction as the predictor of site usage.

## f) Check the Diagnostics

```{r}
#| label: residuals for model 2

par(mfrow = c(2,2))
plot(model2)

```

## g) Visualize Model Predictions

```{r}
#| label: prediction


preds <- ggpredict(model2,
                   terms = c("vo_nest"))
preds_dist <- ggpredict(model3,
                        terms = c("edgedist"))

```

```{r}
#| label: model

ggplot(data = preds, # set up using predictions data frame
       aes(x = x, # set x and y
           y = predicted)) +
  coord_cartesian(ylim = c(0,1)) + # set y limits because values above 1 or below 0 do not make biological or mathematical sense
  geom_smooth(method = "lm",# add prediction line
              color = "deeppink4", # add custom color
              aes(x = x, # set x and y again 
                  y = predicted)) +
  geom_ribbon(alpha = 0.5, # add confidence interval, make transparent
              fill = "pink",# add custom color
              aes(x = x, # add x and y
                  y = predicted, 
                  ymin = conf.low,# add lower interval
                  ymax = conf.high)) +# add upper interval
  geom_jitter(data = mopl_clean, # use original mopl_clean data frame for overlay
              height = 0.03,# set narrower height
              width = 0, # do not jitter width
              alpha = 0.7,# make transparent
              shape = 21,# change shape to open circles
              aes(x = vo_nest, # set x and y axis
                  y = as.numeric(used_bin),# allow plotting of discrete variable over continuous predictions
                  color = used_bin)) +# color by used vs unused site
  scale_color_manual( 
    name = "Actual Site Occupancy",# name legend
    labels = c("Unoccupied", "Occupied"),# change labels from 0 & 1 to unoccupied and occupied
    values = c("brown4","darkolivegreen" # set custom colors
  )) + 
  labs(title = "Visual Obstruction Decreases Plover Nest Selection Probability",# add title
       x = "Visual Obstruction Index", # label x axis
       y = "Predicted Nest Site Occupancy Probability") + # label y axis
  theme_classic() + # change theme 
  theme(plot.title = element_text(hjust = 0.5)) + # move plot title to center
  theme(text = element_text(family = "Times New Roman",
                            color = "#1D0200",
                            size = 16)) + # change font, color,and size
  theme(axis.text = element_text(color = "#1D0200")) # match axis color to text color

```

## h) Caption For Figure

Figure 1. The figure above displays the predicted probability of a prairie dog colony site being selected for use by mountain plovers based on visual obstruction index using a linear regression model. The dark purple line is the predicted probability of selection from 0 to 1 and the pink overlay represents the 95% confidence interval. The green open circles are individual occupied sites and the brown open circles are individual unoccupied sites (the underlying data points have a binary discrete y value (0 and 1 representing unoccupied and occupied) and are underlayed to show difference in visual obstruction index between sites, any jittering around 1 or 0 values is purely for easier visualization). (data obtained from: Duchardt, Courtney J; Beck, Jeffrey L; Augustine, David J (2020). Mountain plover habitat selection and nest survival in relation to weather variability and spacial attributes of black tailed prairie dog disturbance)

## i) Calculate Model Predictions

```{r}

ft <- flextable(as.data.frame(preds)) # create flextable

ft <- colformat_double(ft, # round digits
                       digits = 2)
ft <- autofit(ft) # autofit table

```

```{r}

ft

```

According to the model, the predicted probability of Mountain Plover nest site occupancy at 0 on the index of visual obstruction is 0.66

## j) Interpret Results

When there is no visual obstruction, the probability of nest site occupancy is 0.66. For each 1 index increase in visual obstruction, the probability of nest site occupancy decreases by 0.07 +/- 0.01 (linear model: F(1,274) = 46.42, p \< 0.001, $\alpha$ = 0.05, adjusted R\^2 = 0.14). There is not sufficient evidence to suggest any relationship between distance from prairie colony edge and probability of nest site occupancy (linear model: F(1,274) = 0.5227, p = 0.47, $\alpha$ = 0.05, adjusted R\^2 = -0.0017). The relationship between visual obscurity and Mountain Plover nest site occupancy could be explained by the necessity to avoid predation. When visual obscurity is high, it is difficult to see predators from a distance. Having the clearest view of your surroundings increases survivability, making sites with the best visibility more desirable for Mountain Plovers.

# Problem 4

## a) Comparing Visualizations

Compared to my exploratory visualization, my affective visualization is much more visually interesting and would help a viewer understand the Minecraft apple drop rate much more than my initial visualization did. However, my exploratory visualization does a better job of showing how close the apple drop rates between leaf breaking methods were compared to my affective visualization, which was the purpose of my experiment. Both visualizations show apple drop rate, but my exploratory visualization focuses on the comparison of the precise drop rate per leaf block between methods, while my affective visualization focuses on the less precise drop rate per tree. The overall pattern of apple drop rates not changing based on leaf breaking method stayed the same regardless of how it was visualized.

## b) Designing an Analysis

I propose a Mann-Whitney U test to compare the difference in median apple drops between decayed and punched leaves. The predictor variable (break type) is a categorical variable, while the response (apple drops) is a discrete ordinal numerical variable.

## c) Check Assumptions and Run Analysis

### Assumptions:

```{r}

ggplot(data = apple,# setting up plot with apple data frame
       aes(x = apples, # set x axis
           fill = break_type)) + # fill by break type 
  geom_bar(position = "dodge", # prevent stacking of bars
           color = "#1D0200") + # add outline
  scale_fill_manual( # setting custom colors
    values = c(
      "decay" = "darkolivegreen",
      "punch" = "tomato4"
    )
  ) +
  labs(x = "Apples", # setting labels
       y = "Count",
       title = "Similar Distributions Between Break Type",
       fill = "Break Type") +
  theme_minimal() + # change theme
  theme(plot.title = element_text(hjust = 0.5)) + # move plot title to center
  theme(text = element_text(family = "Times New Roman",
                            color = "#1D0200",
                            size = 16)) + # change font,color and size
  theme(axis.text = element_text(color = "#1D0200")) # match axis color to text color
  
```

Distributions for both variables are extremely similar in shape.

### Analysis:

```{r}

wilcox.test( # run wilcox test to compare medians
  apples ~ break_type,
  data = apple
)
cliffs_delta( # run cliffs delta to find effect size
  apples ~ break_type,
  data = apple
)

```

## d) Visualization

```{r}

ggplot(data = apple, # set up plot with apple data frame
       aes(x = break_type, # set y and x axis
           y = apples,
           fill = break_type)) + # color by break type
  geom_boxplot(varwidth = TRUE,# set width
               linewidth = 1, # increase line width
               linetype = "dashed",
               color = "#1D0200") + # make lines dashed to see median better
  coord_flip() + # flip boxplot horizontally
  scale_fill_manual(values = c( # set custom colors
    "decay" = "darkolivegreen",
    "punch" = "tomato4"
  )) +
  labs(x = "Break Type", # set labels
       y = "Apples Dropped Per Tree",
       title = "No Difference in Median Apple Drops Per Tree",
       fill = "Break Type") +  
  theme_minimal() + # change theme
  theme(legend.position = "none") + # remove legend
  theme(plot.title = element_text(hjust = 0.5)) + # move plot title to center
  theme(text = element_text(family = "Times New Roman",
                            color = "#1D0200",
                            size = 16)) + # change font, color, and size 
  theme(axis.text = element_text(color = "#1D0200")) # match axis color to text color

```

## e) Caption

Figure 2. The figure above compares the median apples dropped per tree between break types. The red boxplot represents trees that had their leaves broken by the player and the green boxplot represents trees with leaves that naturally decayed after log removal. The thicker dashed line is the median (0) and there is only one quartile (1) as it is not possible to have less than 0 apples.

## f) Results

I found almost no difference (Cliff's $\delta$ = -0.01) in median apple drop rate between punched (n = 200) and decayed trees (n = 200) in Minecraft (Mann-Whitney U test, W = 19771, p-value = 0.80). Both trees had a median of 0, interquartile range of 1, upper quartile of 1, and a maximum of 2 apples drop.
