---
title: "ENVS193DS-code"
format: html
editor: visual
author: "Valentina Caro"
date: "03-13-2026"
execute: 
  message: false
  warning: false
---

## Setup

```{r}
#| label: reading in packages
#| message: false

library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # organization
library(readxl) # reading excel
library(scales) # modifying axis labels
library(ggeffects) # model predictions
library(MuMIn) # model selection
library(lterdatasampler) # dataset from LTER
library(extrafont) # adds more fonts
library(showtext) # allow changing font

kelp <- read.csv(
  here("data",
       "Annual_Kelp_All_Years_20250903.csv")
) # creating kelp dataframe

mopl <- read.csv(
  here("data",
       "MOPL_nest-site_selection_Duchardt_et_al._2019.csv")
)

```

HTML Link: <https://valecarov.github.io/ENVS-193DS_winter_2026_final/code/193ds-code.html>

# Problem 1

## a) Transparent Statistical Methods

In part 1, my coworkers used a linear regression to test whether precipitation predicts flooded wetland area. In part 2, my coworkers used a Kruskal-Wallis test to determine whether the median flooded wetland area differs across 5 water year classifications.

## b) Figure Needed

An appropriate figure to accompany part 2 of the report would be a box and jitter plot to show the median and overall spread across the water year classifications. The x-axis/group should be water year classification (wet, above normal, below normal, dry, and critical drought), and the y axis should be flooded wetland area in m\^2. Each classification should have its own custom color to ensure data is clearly displayed.

## c) Suggestions for Rewriting

Based on our model predictions, we found that precipitation does not predict flooded wetland area (m\^2)(Linear Regression, F(1, df) = f-statisic, R^2^ = R\^2 value, p = 0.11, $\alpha$ = significance level). We did not find a difference ( $\eta^2$ = effect size) in median flooded wetland area (m^2)^ between water year classifications (wet, above normal, below normal, cry, critical drought) (Kruskal-Wallis test, $\chi^2$ (df) = H-statistic, p = 0.12, $\alpha$ = significance level).

## d) Interpretation

One additional variable that could impact the flooded wetland area of the San Joaquin River Delta is water management. As there is not sufficient evidence of precipitation having an effect on flooded wetland area in this study, it is likely that the management of the water is a stronger predictor than the amount of water itself. This could be tested by measuring the flooded wetland area under different water management strategies to determine which strategies may increase flooded wetland area, maximizing waterfowl breeding success.

# Problem 2

## a) Cleaning and Summarizing

```{r}
#| label: cleaning & summarizing

kelp_clean <- kelp |> 
  clean_names() |> # clean the names
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> # filter for sites IVEE, NAPL, & CARP
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> # exclude specific date, transect, and quad values
  mutate(site_name = case_when( # create column for site names from site, using full name rather than abbreviation
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) |> 
  mutate(site_name = as_factor(site_name),
       site_name = fct_relevel(site_name,
                               "Naples", "Isla Vista", "Carpinteria")) |> # make site name a factor and set order
  group_by(site_name, year) |> # group by site name and year
  summarize(mean_fronds = mean(fronds)) |> # calculate the mean frond size
  ungroup() # undo group_by function

```

```{r}
#| label: check structure

str(kelp_clean) # check structure of the data frame

```

```{r}
#| label: display

slice_sample(kelp_clean,
             n = 5) # display 5 random rows from data frame

```

## b) Understanding the Data

Data from 09-22-2000 in transect 6 in quad 40 was excluded from kelp_clean due to a missing data sheet causing invalid values for the number of fronds and the diameter of the holdfast. The number and size (number of fronds and diameter of the holdfast) of kelp in this line of the table are both -99999. (Abstract section under Summary tab of data page)

## c) Visualizing the Data

```{r}
#| label: add font

font_add(family = "Times New Roman", regular = "times.ttf") # adds Times New Roman font
showtext_auto() # allows font to be used 

```

```{r}
#| label: line graph

ggplot(data = kelp_clean,# use kelp_clean data frame
       aes(x = year, # set x and y axis
           y = mean_fronds,
           color = site_name)) +
  geom_line() + # adding line graph
  geom_point() + # adding points
  labs(title= "Sites vary in mean kelp fronds per year", # setting titles
       x = "Year",
       y = "Mean kelp fronds per year",
       color = "Site") +
  scale_color_manual(values = c( # set custom colors
    "Naples" = "darksalmon",
    "Isla Vista" = "palevioletred",
    "Carpinteria" = "deeppink4"
  )) +
  theme_minimal() + # change theme 
  theme(plot.title = element_text(hjust = 0.5)) + # move plot title to center
  theme(text = element_text(family = "Times New Roman", # change font and size
                            size = 16)) 
  
```

## d) Interpretation

The site with the most variability in mean kelp fronds per year was the Naples site, with values as low as about 4 fronds up to nearly 40 fronds, visualized by the huge spikes in the graph above. The Carpinteria site had the least variability in mean kelp fronds per year, ranging from about 3 fronds to 17 fronds as seen by the much smaller spikes.

# Problem 3

## a) Response Variable

In this dataset, the 0s and 1s in the binshrub and binsubshrub columns indicate whether there is shrub or sub-shrub cover in a nest-site. A 0 indicates no shrub or sub-shrub cover and a 1 indicates there is shrub or sub-shrub cover in the nest-site.

## b) Purpose of Study

### Visual Obstruction:

Visual obstruction is a continuous variable measured at the nest, and at 5 m and 10 m from the nest. As visual obstruction increases, the probability of Mountain Plover nest site use would decrease because mountain plovers have a positive response to sites with a higher percentage of bare ground (the least visual obstruction) and a negative response to increasing maximum vegetation height, and presence of shrub cover and cactus cover (higher visual obstruction) according to the nest-selection subsection under the results section.

### Distance to Prairie Dog Colony Edge:

Distance to prairie dog colony edge is a continuous variable. As distance to the colony edge increases, the probability of a Mountain Plover nest site would decrease. According to the 2nd and 3rd paragraphs in the Discussion section of the paper, this may be due to complementary resources outside of the colony. It may be easier for adult plovers to travel with chicks if they are closer to the edge of the colony rather than at the center. Additionally, mountain plover density was highest in colonies that were 100-500 ha in size and low in larger colonies according to Figure 2 in the Results section, which could be related to distance from the edge as plovers would have to travel much further distances in very large colonies. I believe mountain plovers prefer nest sites that offer protection while still allowing for easy travel to outside resources, causing density to be the highest near the edges of moderately sized colonies.

## c) Table of Models

\| Model Number \| Visual Obstruction \| Distance to Prairie Dog Colony \| Model Description \|

\|:-----------------:\|:---------------------:\|:----------------------------------:\|:--------------------:\|

\|:--------0--------:\|:---------------------:\|:----------------------------------:\|:--------null--------:\|

\|:--------1--------:\|:---------x-----------:\|:----------------x-----------------:\|:--------all---------:\|

\|:--------2--------:\|:---------x-----------:\|:----------------------------------:\|:--------vo---------:\|

\|:--------3--------:\|:---------------------:\|:----------------x-----------------:\|:-----edgedist-----:\|

## d) Run the Models

```{r}
#| label: cleaning

mopl_clean <- mopl |> # create new dataframe
  clean_names() |> # clean names
  mutate( used_bin = case_when( # create new column used_bin from id column
    id == "used" ~ "1",
    id == "available" ~ "0"
  )) |> 
  select(used_bin, edgedist, vo_nest) # select variables

```

```{r}
#| label: run models
#| include: false

# model 0: null model
model0 <- lm(used_bin ~ 1,
             data = mopl_clean)
# model 1: all predictors
model1 <- lm(used_bin ~ vo_nest + edgedist,
             data = mopl_clean)
# model 2: visual obstruction
model2 <- lm(used_bin ~ vo_nest,
             data = mopl_clean)
# model 3: distance to edge of colony
model3 <- lm(used_bin ~ edgedist,
             data = mopl_clean)

```
